---
title: "Non-Parametric Methods: Kaplan-Meier Estimator"
author: "José Jorge Martínez de la Cruz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Kaplan Meier
The Kaplan-Meier estimator is a non-parametric statistic used to estimate the survival function from lifetime data. In this session, we will reconstruct the estimator manually using dplyr to understand its mechanics before using the standard survival package functions

```{r}
library(dplyr)
library(survival)
library(ggplot2)
```
We use data from a clinical trial comparing a treatment group against a placebo group.

```{r}
# Treatment group: times and event status (1 = event, 0 = censored)
times_treat <- c(6,6,6, 7, 10, 13, 16, 22, 23, 6, 9, 10, 11, 17, 19, 20, 25, 32, 32, 34, 35)
status_treat  <- c(rep(1, 9), rep(0, 12))

# Placebo group (no censoring)
times_plac <- c(1,1, 2,2, 3, 4,4, 5,5, 8,8,8,8, 11,11, 12,12, 15, 17, 22, 23)
status_plac  <- rep(1, 21) 

# Combine into a single data frame
df <- data.frame(
  time = c(times_treat, times_plac),
  status = c(status_treat, status_plac),
  group  = rep(c("Treatment", "Placebo"), each = 21)
)
```

The survival probability at time $t$ is calculated as the product of conditional survival probabilities:
$$\hat{S}(t) = \prod_{t_i \leq t} \left(1 - \frac{d_i}{n_i}\right)$$

Where $d_i$ is the number of events at time $t_i$ and $n_i$ is the number of individuals at risk just before time $t_i$.

```{r}
# Step-by-step manual estimation using dplyr
KM_manual <- df %>%
  group_by(group, time) %>%
  summarise(
    d_j = sum(status == 1),       # Number of events at time j
    c_j = sum(status == 0),       # Number of censored observations at time j
    .groups = "drop"
  ) %>%
  group_by(group) %>%
  arrange(group, time) %>%
  mutate(
    # Number at risk (n_j): 21 minus cumulative sum of (events + censored)
    n_j = 21 - lag(cumsum(d_j + c_j), default = 0),
    # Interval survival probability
    interval_survival = 1 - (d_j / n_j),
    # Cumulative product for S(t)
    S_t = cumprod(interval_survival)
  ) %>%
  # Filter to show only event times
  filter(d_j > 0) %>%
  select(group, time, n_j, d_j, S_t)

print(KM_manual)
```

Using the survfit object, we can create a step-plot comparison. Censored observations are traditionally marked with a cross (+).

```{r}
# Standard fit using the survival package
km_fit <- survfit(Surv(time = df$time, event = df$status) ~ group, data = df)
summary(km_fit)
```

```{r}

# Prepare data for ggplot
df_plot <- data.frame(
  time = km_fit$time,
  survival = km_fit$surv,
  group = rep(names(km_fit$strata), km_fit$strata),
  censored = km_fit$n.censor
)

# Clean group names
df_plot$group <- gsub("group=", "", df_plot$group)

# Add starting point (time 0, survival 100%)
df_start <- data.frame(
  time = c(0, 0),
  survival = c(1, 1),
  group = c("Treatment", "Placebo"),
  censored = c(0, 0)
)
df_plot <- rbind(df_start, df_plot)

# Final Visualization
ggplot(df_plot, aes(x = time, y = survival, color = group)) +
  geom_step(linewidth = 1) + 
  geom_point(data = subset(df_plot, censored > 0), 
             aes(x = time, y = survival), 
             shape = 3, size = 3, stroke = 1, show.legend = FALSE) +
  scale_color_manual(values = c("Placebo" = "green", "Treatment" = "blue")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 40, by = 5)) +
  labs(
    title = "Kaplan-Meier Survival Estimates",
    subtitle = "Comparison between Treatment and Placebo cohorts",
    x = "Weeks",
    y = "S(t)",
    color = "Cohort"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

