---
title: "Parametric Survival Models and Kaplan-Meier Comparison"
author: "Jorge de la Cruz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to Parametric Modeling

In survival analysis, parametric models assume that the survival time $T$ follows a specific probability distribution. In this session, we explore the Exponential, Weibull, and Log-Normal distributions using lung cancer data.

```{r}
library(survival)
```
## Data Loading and Preprocessing

We use the NCCTG Lung Cancer dataset to model survival times.

```{r}
# Ensure the data file is in your working directory
datos <- read.table(file = "NCCTGLungCancer.txt", header = TRUE)
# For demonstration, we'll use the built-in 'lung' dataset from survival package
data(lung)
datos <- lung
head(datos)  
str(datos)
```

## Fitting Parametric Models

### Exponential Model
The Exponential distribution assumes a constant hazard rate $\lambda$.

```{r}
modelo_exponencial <- survreg(Surv(time, status) ~ 1, 
                              data = datos, 
                              dist = "exponential")

summary(modelo_exponencial)

# In survreg, the intercept is log(1/lambda)
lambda_estimado <- exp(-coef(modelo_exponencial)[1])

# Plotting the density function f(t)
tiempos <- seq(0, max(datos$time, na.rm = TRUE), length.out = 500)

hist(datos$time, 
     freq = FALSE, 
     main = "Histogram vs. Exponential Density",
     xlab = "Time (t)", 
     ylab = "f(t)",
     col = "lightgray", 
     border = "white",
     ylim = c(0, 0.005))

densidad_exp <- lambda_estimado * exp(-lambda_estimado * tiempos)
lines(tiempos, densidad_exp, col = "blue", lwd = 3, lty = 1)

legend("topright", 
       legend = c("Histogram", "Exponential f(t)"),
       fill = c("lightgray", NA),
       col = c(NA, "blue"),
       border = c("white", NA),
       lwd = c(NA, 3),
       bty = "n")
```
### Weibull Model

The Weibull distribution is more flexible as it allows for increasing or decreasing hazard rates over time.

```{r}
modelo_weibull <- survreg(Surv(time, status) ~ 1, 
                          data = datos, 
                          dist = "weibull")

# Scale (lambda) and Shape (k) parameters
escala_weibull <- exp(coef(modelo_weibull)[1]) 
forma_weibull  <- 1 / modelo_weibull$scale     

hist(datos$time, 
     freq = FALSE, 
     main = "Histogram vs. Weibull Density",
     xlab = "Time (t)", 
     ylab = "Density",
     col = "lightgray", 
     border = "white",
     ylim = c(0, 0.0035)) 

densidad_weibull <- dweibull(tiempos, 
                             shape = forma_weibull, 
                             scale = escala_weibull)

lines(tiempos, densidad_weibull, col = "purple", lwd = 3)         

legend("topright", 
       legend = c("Histogram", "Weibull Density"),
       fill = c("lightgray", NA),
       col = c(NA, "purple"),
       border = c("white", NA),
       lwd = c(NA, 3),
       bty = "n")
```
### Log-Normal Model
The Log-Normal model is often used when the hazard rate initially increases and then decreases.

```{r}
modelo_lognormal <- survreg(Surv(time, status) ~ 1, 
                            data = datos, 
                            dist = "lognormal")

mu_lognormal <- coef(modelo_lognormal)[1]
sigma_lognormal <- modelo_lognormal$scale

hist(datos$time, 
     freq = FALSE, 
     main = "Histogram vs. Log-Normal Density",
     xlab = "Time (t)", 
     ylab = "Density",
     col = "lightgray", 
     border = "white",
     ylim = c(0, 0.0035)) 

densidad_lognormal <- dlnorm(tiempos, 
                             meanlog = mu_lognormal, 
                             sdlog = sigma_lognormal)

lines(tiempos, densidad_lognormal, col = "darkorange", lwd = 3)

legend("topright", 
       legend = c("Histogram", "Log-Normal Density"),
       fill = c("lightgray", NA),
       col = c(NA, "darkorange"),
       border = c("white", NA),
       lwd = c(NA, 3),
       bty = "n")
```

## Kaplan-Meier Comparison

Finally, we compare the survival functions $S(t)$ of all parametric models against the non-parametric Kaplan-Meier estimate to see which one fits the data best.

```{r}
# Kaplan-Meier fit
km_fit <- survfit(Surv(time, status) ~ 1, data = datos)

plot(km_fit, 
     main = "Model Comparison: S(t)",
     xlab = "Time (t)", 
     ylab = "Survival Probability S(t)",
     col = "black", 
     lwd = 3,
     conf.int = FALSE) 

# Calculate Survival Functions for each model
S_exp <- exp(-lambda_estimado * tiempos)
S_weibull <- pweibull(tiempos, shape = forma_weibull, scale = escala_weibull, lower.tail = FALSE)
S_lognorm <- plnorm(tiempos, meanlog = mu_lognormal, sdlog = sigma_lognormal, lower.tail = FALSE)

# Add lines to the plot
lines(tiempos, S_exp, col = "red", lwd = 2, lty = 2)
lines(tiempos, S_weibull, col = "purple", lwd = 2, lty = 2)
lines(tiempos, S_lognorm, col = "darkorange", lwd = 2, lty = 1)

legend("topright", 
       legend = c("Kaplan-Meier (Non-parametric)", 
                  "Exponential", 
                  "Weibull", 
                  "Log-Normal"),
       col = c("black", "red", "purple", "darkorange"),
       lwd = c(3, 2, 2, 2),
       lty = c(1, 2, 2, 1),
       bty = "n",
       cex = 0.8)
```

