---
title: "Variable Selection and Linearity Diagnostics in Cox Models"
author: "José Jorge Martínez de la Cruz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(collett)
```

# ### Example 1
 This example analyzes the survival time of breast cancer patients, classified according to the staining result of their tumors. The indicator variable takes the value of 1 for positive staining and 0 for negative. 

```{r}
data(bcancer) # The breast cancer database is loaded


bcancer$staining <- bcancer$stain - 1 

# The Cox model is fitted
modelo_1 <- coxph(Surv(time, status) ~ staining, data = bcancer, ties = "breslow")

# results
summary(modelo_1)

```

# ### Example 2
In this second case, the survival of 48 patients is modeled using seven explanatory variables: Age, Sex, Bun, Ca, Hb, Pcells, and Protein.

```{r}
data(myeloma) # The myeloma database is loaded


modelo_2 <- coxph(Surv(time, status) ~ age + sex + bun + ca + hb + pcells + protein, 
                         data = myeloma)

# The model is fitted without the 'bun' variable
modelo_2_1<- coxph(Surv(time, status) ~ age + sex + ca + hb + pcells + protein, 
                         data = myeloma)

# Both models are compared
summary(modelo_2)

summary(modelo_2_1)

```

```{r}

get_2logL <- function(modelo) {
  return(-2 * modelo$loglik[2]) 
}

```

# Example 3. Survival of multiple myeloma patients

```{r}
modelo_3_nulo <- coxph(Surv(time, status) ~ 1, data = myeloma)

# Models with one variable
modelo_3_bun <- coxph(Surv(time, status) ~ bun, data = myeloma)

modelo_3_age <- coxph(Surv(time, status) ~ age, data = myeloma)

modelo_3_sex <- coxph(Surv(time, status) ~ sex, data = myeloma)

modelo_3_ca <- coxph(Surv(time, status) ~ ca, data = myeloma)

modelo_3_pcells <- coxph(Surv(time, status) ~ pcells, data = myeloma)

modelo_3_hb  <- coxph(Surv(time, status) ~ hb, data = myeloma)

modelo_3_protein <- coxph(Surv(time, status) ~ protein, data = myeloma)

#-2loglik
cat("Null -2 log L:", -2 * modelo_3_nulo$loglik[1], "\n")
cat("Age -2 log L:", get_2logL(modelo_3_age), "\n")
cat("Sex -2 log L:", get_2logL(modelo_3_sex), "\n")
cat("BUN -2 log L:", get_2logL(modelo_3_bun), "\n")
cat("Ca -2 log L:", get_2logL(modelo_3_ca), "\n")
cat("Hb -2 log L:", get_2logL(modelo_3_hb), "\n")
cat("Pcells -2 log L:", get_2logL(modelo_3_pcells), "\n")
cat("Protein -2 log L:", get_2logL(modelo_3_protein), "\n")

```

```{r}
# Model summaries to see the p-values of each variable
summary(modelo_3_bun)

summary(modelo_3_hb)

summary(modelo_3_protein)

```


```{r}

# Combined models
mod_3_bun_hb_protein <- coxph(Surv(time, status) ~ bun + hb+ protein, data = myeloma)
mod_3_bun_hb <- coxph(Surv(time, status) ~ bun + hb, data = myeloma)
mod_3_bun_protein <- coxph(Surv(time, status) ~ bun + protein, data = myeloma)
mod_3_protein_hb <- coxph(Surv(time, status) ~ hb + protein, data = myeloma)

cat("Hb + bun + protein -2 log L:", get_2logL(mod_3_bun_hb_protein), "\n")
cat("Hb + bun -2 log L:", get_2logL(mod_3_bun_hb), "\n")
cat("protein + bun -2 log L:", get_2logL(mod_3_bun_protein), "\n")
cat("Hb + protein -2 log L:", get_2logL(mod_3_protein_hb), "\n")


```

```{r}
# Variables that were previously not significant are added
mod_3_bun_hb_age <- coxph(Surv(time, status) ~ bun + hb+age, data = myeloma)
mod_3_bun_hb_sex<- coxph(Surv(time, status) ~ bun + hb + sex, data = myeloma)
mod_3_bun_hb_ca<- coxph(Surv(time, status) ~ bun + hb + ca, data = myeloma)
mod_3_bun_hb_pcells <- coxph(Surv(time, status) ~ bun + hb + pcells, data = myeloma)

# Print the -2loglik statistics
cat("Hb + BUN + Age -2 log L:", get_2logL(mod_3_bun_hb_age), "\n")
cat("Hb + BUN + Sex -2 log L:", get_2logL(mod_3_bun_hb_sex), "\n")
cat("Hb + BUN + Ca -2 log L:", get_2logL(mod_3_bun_hb_ca), "\n")
cat("Hb + BUN + Pcells -2 log L:", get_2logL(mod_3_bun_hb_pcells), "\n")

```

```{r}
summary(mod_3_bun_hb_sex)
```


# Example 4: Treatment, Prostatic cancer 

```{r}
data(prostatic)
prostatic$treatment <- prostatic$treatment-1

library(survival)

# Fit all possible models


# 0 variables (Null model)
mod_none <- coxph(Surv(time, status) ~ 1, data = prostatic)

# 1 variable
mod_age   <- coxph(Surv(time, status) ~ age, data = prostatic)
mod_shb   <- coxph(Surv(time, status) ~ shb, data = prostatic)
mod_size  <- coxph(Surv(time, status) ~ size, data = prostatic)
mod_index <- coxph(Surv(time, status) ~ index, data = prostatic)

# 2 variables
mod_age_shb   <- coxph(Surv(time, status) ~ age + shb, data = prostatic)
mod_age_size  <- coxph(Surv(time, status) ~ age + size, data = prostatic)
mod_age_index <- coxph(Surv(time, status) ~ age + index, data = prostatic)
mod_shb_size  <- coxph(Surv(time, status) ~ shb + size, data = prostatic)
mod_shb_index <- coxph(Surv(time, status) ~ shb + index, data = prostatic)
mod_size_index<- coxph(Surv(time, status) ~ size + index, data = prostatic)

# 3 variables
mod_age_shb_size   <- coxph(Surv(time, status) ~ age + shb + size, data = prostatic)
mod_age_shb_index  <- coxph(Surv(time, status) ~ age + shb + index, data = prostatic)
mod_age_size_index <- coxph(Surv(time, status) ~ age + size + index, data = prostatic)
mod_shb_size_index <- coxph(Surv(time, status) ~ shb + size + index, data = prostatic)

# 4 variables
mod_todos <- coxph(Surv(time, status) ~ age + shb + size + index, data = prostatic)


cat("Model none AIC:", AIC(mod_none), "\n")

cat("\n--- Models with 1 variable ---\n")
cat("Model Age AIC:", AIC(mod_age), "\n")
cat("Model Shb AIC:", AIC(mod_shb), "\n")
cat("Model Size AIC:", AIC(mod_size), "\n")
cat("Model Index AIC:", AIC(mod_index), "\n")

cat("\n--- Models with 2 variables ---\n")
cat("Model Age + Shb AIC:", AIC(mod_age_shb), "\n")
cat("Model Age + Size AIC:", AIC(mod_age_size), "\n")
cat("Model Age + Index AIC:", AIC(mod_age_index), "\n")
cat("Model Shb + Size AIC:", AIC(mod_shb_size), "\n")
cat("Model Shb + Index AIC:", AIC(mod_shb_index), "\n")
cat("Model Size + Index AIC:", AIC(mod_size_index), "\n")

cat("\n--- Models with 3 variables ---\n")
cat("Model Age + Shb + Size AIC:", AIC(mod_age_shb_size), "\n")
cat("Model Age + Shb + Index AIC:", AIC(mod_age_shb_index), "\n")
cat("Model Age + Size + Index AIC:", AIC(mod_age_size_index), "\n")
cat("Model Shb + Size + Index AIC:", AIC(mod_shb_size_index), "\n")

cat("\n--- Model with 4 variables ---\n")
cat("Model Age + Shb + Size + Index AIC:", AIC(mod_todos), "\n")

```


```{r}
# Now we fit the treatment
mod_size_index_trat<- coxph(Surv(time, status) ~ size + index + treatment, data = prostatic)

cat("Model Treatment + Size + Index AIC:", AIC(mod_size_index_trat), "\n")

```


# Preliminary linearity test:

```{r}
# Example 4

cat("\nModel with Hb:", get_2logL(mod_3_bun_hb), "\n")

#  Test adding a quadratic term
myeloma$hb_cuadrado <- myeloma$hb^2
mod_hb_cuad <- coxph(Surv(time, status) ~ bun + hb + hb_cuadrado, data = myeloma)
cat("\nModel with squared Hb -2 log L:", get_2logL(mod_hb_cuad), "\n")

#  Test using categorization (Factors)
myeloma$hb_factor <- cut(myeloma$hb, 
                         breaks = c(-Inf, 7, 10, 13, Inf), 
                         labels = c("1", "2", "3", "4"))

mod_hb_factor <- coxph(Surv(time, status) ~ bun + hb_factor, data = myeloma)
cat("Model with Hb as Factor -2 log L:", get_2logL(mod_hb_factor), "\n")

```

The AIC does not change when we introduce a quadratic term, but it does when we introduce the factor variable, so we calculate the p-value of this difference:

```{r}
# Compare the model with Hb as a continuous variable against the model with Hb as a factor
anova(mod_3_bun_hb, mod_hb_factor)
```



 
 