---
title: 'Workshop: Generalized Linear Models (GLM)'
author: "José Jorge Martínez de la Cruz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to Generalized Linear Models (GLM)
Generalized Linear Models (GLMs) extend ordinary linear regression by allowing the response variable to have an error distribution other than a normal distribution. In this workshop, we cover Logistic Regression for binary outcomes and Poisson Regression for count data

```{r}
# Load necessary libraries
library(tidyverse)
library(broom)
library(margins)
library(pscl)

set.seed(236)
```
## Logistic Regression: Titanic Dataset
We use the Titanic dataset to model the probability of survival based on socio-economic class, sex, and age.

```{r}
# Load and prepare the Titanic dataset
data("Titanic")
titanic_df <- as.data.frame(Titanic)

# Expand frequencies to create a tidy dataset
titanic_expanded <- titanic_df[rep(1:nrow(titanic_df), titanic_df$Freq), ]
titanic_expanded$Freq <- NULL

# Create binary survival variable
titanic_expanded$Survived <- ifelse(titanic_expanded$Survived == "Yes", 1, 0)

str(titanic_expanded)
```
### Model fit and interpretation

```{r}
# Fit Logistic Model
logit_model <- glm(Survived ~ Class + Sex + Age,
                   data = titanic_expanded,
                   family = binomial(link = "logit"))

summary(logit_model)

# Calculate Odds Ratios: exp(beta)
odds_ratios <- exp(coef(logit_model))
print(odds_ratios)

# Calculate Marginal Effects
# These represent the change in probability for a unit change in the predictor
logit_margins <- margins(logit_model)
summary(logit_margins)
```
## Poisson Regression: Count Data

We analyze the warpbreaks dataset, which records the number of breaks in yarn during weaving.

```{r}
data("warpbreaks")
summary(warpbreaks)
```

### Model fit and interpretation
```{r}
# Fit Poisson Model
poisson_model <- glm(breaks ~ wool + tension,
                     data = warpbreaks,
                     family = poisson(link = "log"))

summary(poisson_model)

# Calculate Incidence Rate Ratios (IRR)
irr_poisson <- exp(coef(poisson_model))
print(irr_poisson)

# Overdispersion Diagnostic (Ratio of Pearson chi-square to degrees of freedom)
# If dispersion > 1, the variance exceeds the mean, violating Poisson assumptions.
dispersion_ratio <- sum(residuals(poisson_model, type = "pearson")^2) /
                    poisson_model$df.residual
cat("Dispersion Ratio:", dispersion_ratio, "\n")
```

When overdispersion is present (as seen above), we use a Quasi-Poisson model to adjust the standard errors

```{r}
# Fit Quasi-Poisson Model
quasi_model <- glm(breaks ~ wool + tension,
                   data = warpbreaks,
                   family = quasipoisson(link = "log"))

summary(quasi_model)
```
Note on Quasi-Poisson: We do not aim to eliminate overdispersion but to estimate it through the dispersion parameter $\phi$. The key comparison is that the Standard Errors (SE) in the Quasi-model should be larger than in the standard Poisson model, leading to more reliable p-values.
